Propositions as Sessions, Mechanically

### 2015-04-08 paper/formalisation.tex
Added a section discussing the failed attempt at cut elimination.

% Description of the typing rules...

Figure~\ref{fig:cll} presents the process calculus CP in LNCQ form with some
tedious details relating to environment uniqueness and ordering (see below)
elided. In other words, bound variables are represented using de Brujin
indices and free variables are represented using an abstract type provided by
the Metatheory library. The rules with binders reflect this by specifying the
type of a bound variable in the constructor rather than a name. Cofinite
quantification is performed in the hypotheses to these rules by excluding a
finite set of names from opening the subprocesses where opening a process $P$
with name $x$ is denoted $P^x$. Note that the rules for polymorphism,
$\exists$ and $\forall$, were not included in the final Coq development since
GV (\S~\ref{sec:gv}) does not support polymorphism making the rules
unnecessary for the translation. Further, as noted by \citeauthor{Lee:2012},
support for polymorphism in type variables creates an explosion of the
infrastructure requiring additional of opening and substituting definitions
and lemmas. They appear in the Figure simply for the sake of completeness.

Formalising the process calculus CP provided many challenges for the
Metatheory library since to my knowledge, the library has never been applied
to a process calculus before, and the treatment of names, binding and
substitution differ to that present in $\lambda$-calculi. For instance, in the
$\stlc$ one typically does not work up to permutation of binders, whereas in a
process calculus these kinds of structural equivalences are
standard. Additionally, substitution and opening are only defined for
names. In other words, one cannot substitute an arbitrary process term for a
name, unlike in the $\lambda$-calculus. Thus frameworks which assume
$\lambda$-calculi style operations are not immediately amenable to the CP
setting~\cite{Lee:2012}.

Ordering in environments is ignored so the development required operations for
handling permutations of environments. The Metatheory library does not
currently support permutating environments, however some basic lemmas were
added as a result of this work, and custom tactics were developed to
automatically solve certain permutation goals via simple proof search. Even
so, there are still many cases where sequences of primitive lemmas, such as
those handling append, commutativity and transitivity, need to be applied
manually. A lot of these cases are similar but there would not appear to be an
advantage in developing a tactic to do the work. Doing so would create a
dependency between a hard-to-debug and obscure Ltac definition and the proofs
which utilise it, becoming a burden to change. Further, such a development
philosophy treats Ltac tactics akin to macros, a shorthand for expressing an
oft-repeated sequence of tactics, and it is doubtful much can be gained from
it unless the execution of Ltac tactics can be made more transparent.

@@

The principal cut elimination rules are presented in Figure~\ref{??}. The
reduction rules were complicated to prove due to being composed of multiple
constructor instances and differ significantly to the standard operational
semantics one typically deals with in $\lambda$-calculus formalisms. As a
result, these well-typed reductions were not proved ``in-place'', instead each
rule is proved as a separate lemma. Most proofs followed a pattern but it was
difficult to automate satisfactorily for the reasons stated in the previous
paragraph. However, some tasks could be delegated to a custom tactic, for
example instantiating cofinitely quantified hypotheses with a newly introduced
fresh variable.

Initially, an implicit notion of weakening was chosen so as to be in line with
\citeauthor{Wadler:2014}'s definitions. However, explicit weakening is
required to differentiate the principal cut elimination rules for the various
interactions between servers and clients ($\beta_{!?}$ and $\beta_{!W}$). The
paper presentation does not specify where weakening could occur; it is tacitly
assumed weakening can always be pushed further up the derivation tree to a
base rule e.g. $\Ax\Cut$ rule. Contraction is not supported in the
formalisation due to the difficulty in specifying its principal cut
elimination rule. In order to specify the rule one requires to show the
following:
\begin{gather*}
\tm{\nu A.(!0(A).P \mid ?z[[A]].Q)}
\quad \becomes \quad \\
\tm{\nu A.(!0(A).P \mid \nu A.(!0(A).P \mid \key{contract}_{\Gamma} Q))}
\end{gather*}

@@

Unfortunately, the reduction relation does not have knowledge of the
environment, $\Gamma$, typing process $\tm{P}$. This fact does not impact upon
expressing weakening because only the names (not the types in the bindings)
within the environment are required for weakening. one can compare alms with this version of cp
since it weakening rule corresponds to affine logic

In the empty choice rule there may only be a single binding in the context
(cf. an arbitrary context in \citeauthor{Wadler:2014}'s
presentation~\cite{Wadler:2014}). While this diverges with the presentation of
linear logic it became necessary to do this because otherwise there would be
no relationship between the free variables within a process and its
environment. no there is a clear iff relationship which permits reduction
rules to apply such as i think weakening but it should be any rule which takes
some fvs of proc and infers something about environments from that...otherwise
what it is not at all clear how to encode the reductions in a relation, the
subprocesses environments are created from destructing the environment

@@

more elaborate than first thought. complicated measure.  as it stands the
current definitions of well-typed terms and reduction are not amenable to the
proving of the cut elimination theorem. some method of measure is necessary in
order to develop some notion of termination and also some relationship to the
the the underlying subprocesses additionally there is also the issue of the
representation which includes cases that are not immediately clear how to
prove for instance the commuting conversions are one such example the
reduction is only defined if the logical operator corresponds to a free
variable but induction on the typing judgement leads to two cases one where
the logical operator is applied to the top-level cut variable it is not
immediately clear how to eliminate this from happening a relationship between
the two subprocesses of the cut would need to be stated to prevent them from
differing so that a principal cut reduction rule may be applied yes only then
may the the rule be reduced to a cut-free thing if we had a relationship
between them then there would be some kind of thing that would allow
destruction on one then fixing the other in the case of a bvar equal to zero
\todo{how to proceed?}

>>>>

Figure~\ref{fig:cll} presents the well-typed process relation for CP as an
inductive type in Coq. Each typing rule in CP is represented as a constructor
of the inductive type \coqe$cp_rule$. The inductive type is indexed by a
process (\coqe$proc$) and an environment (\coqe$penv$) indicated by them
appearing after the colon on the first line. \coqe$forall$ quantifies over
hypotheses in a constructor; everything before the comma. Compound hypotheses
are given labels, such as \key{PER}, which is the default name used to
introduce them into the proof context during theorem proving. Some details
such as environment uniqueness have been elided, but notice the use of LNCQ
form within the hypotheses of the rules with binders. These rules exclude a
finite set of names \coqe$L$ from opening the subprocesses where opening a
process \coqe$P$ with name \coqe$x$ is denoted by \coqe$open_proc P x$. Bound
variables are represented using de Bruijn indices and free variables are
represented using an abstract type, \coqe$atom$, provided by the Metatheory
library. The rules with binders specify the type of a bound variable in the
constructor rather than a name. Note that the rules for polymorphism,
$\exists$ and $\forall$, were not included in the final Coq development since
GV (\S~\ref{sec:gv}) does not support polymorphism making the rules
unnecessary for the translation. Further, as noted by \citeauthor{Lee:2012},
support for polymorphism in type variables creates an explosion of the
infrastructure requiring additional of opening and substituting definitions
and lemmas.

Formalising the process calculus CP provided many challenges for the
Metatheory library since to my knowledge, the library has never been applied
to a process calculus before, and the treatment of names, binding and
substitution differ to that present in $\lambda$-calculi. For instance, in the
$\stlc$ one typically does not work up to permutation of binders, whereas in a
process calculus these kinds of structural equivalences are
standard. Additionally, substitution and opening are only defined for
names. In other words, one cannot substitute an arbitrary process term for a
name, unlike in the $\lambda$-calculus. Thus, frameworks which assume
$\lambda$-calculi style operations are not immediately amenable to the CP
setting~\cite{Lee:2012}.

Ordering in environments is ignored so the development required operations for
handling permutations of environments as shown in the \coqe$cp_rule$
definition. The Metatheory library does not currently support permutating
environments, however some basic lemmas were added as a result of this work,
and custom tactics were developed to automatically solve certain permutation
goals via simple proof search. Even so, there are still many cases where
sequences of primitive lemmas, such as those handling append, commutativity
and transitivity, need to be applied manually. A lot of these cases are
similar but there would not appear to be an advantage in developing a tactic
to do the work. Doing so would create a dependency between a hard-to-debug and
obscure Ltac definition and the proofs which utilise it, becoming a burden to
change. Further, such a development philosophy treats Ltac tactics akin to
macros, a shorthand for expressing an oft-repeated sequence of tactics, and it
is doubtful much can be gained from it unless the execution of Ltac tactics
can be made more transparent.

\subsubsection{Subject Reduction Theorem}

The subject reduction theorem can be expressed in Coq as follows:
\begin{coq}
Theorem proc_sub_red:
  forall $\Gamma$ P Q
         (WT: P $\tpvdash_{cp}$ $\Gamma$)
         (RED: P $\becomes$ Q),
    Q $\tpvdash_{cp}$ $\Gamma$.
Proof. ii; gen $\Gamma$; induction RED; subst; eauto. Qed.
\end{coq}

\coqe$Theorem$ or \coqe$Lemma$ indicate the start of a proposition to be
proved (Coq makes no distinction between the two keywords). The statement to
prove is that if \coqe$P$ is a well-typed process in a context $\Gamma$ and
\coqe$P$ reduces to \coqe$Q$ then \coqe$Q$ is well-typed in $\Gamma$. The
proof environment is started using \coqe$Proof.$ and then followed by a
sequence of tactics until all goals have been proven. After execution of
\coqe$Qed.$ the Coq system inspects the validity of the proof object
constructed by the tactics and, if valid, will add the theorem to the
environment.

Some tactics used above are shorthands for commonly combined tactics, for
example \coqe$ii$ repeatedly performs hypothesis introduction then
simplification until no more can be peformed, and \coqe$gen$ $\Gamma$
generalises the goal over the typing environment $\Gamma$.

@@

The \key{RED} hypothesis refers to an inductive type defining the principal
cut reductions and commuting conversions on processes. The inductive cases in
\coqe$proc_sub_red$ were complicated to prove due to the process terms being
composed of multiple constructor instances, differing significantly from the
standard operational semantics typically dealt with in $\lambda$-calculus
formalisms. To mitigate the complexity, the proof for each case was written as
a separate lemma (termed ``reduction lemmas''). The last step in the proof of
\coqe$proc_sub_red$, \coqe$eauto$, solves the goals generated by the induction
using the reduction lemmas.

An example of a reduction lemma is shown in Figure~\ref{fig:principal}. Cut of
server accept against client request involves three constructor instances, the
outer cut and the two dual constructors as the subprocesses. As shown, to
specify this compound process requires the explicit use of de Bruijn
indices. These indices will be replaced in the proof by a sufficiently fresh
name after destructing \key{WT} to obtain the derivations for \coqe$P$ and
\coqe$Q$. Most proofs followed a pattern but it was difficult to automate
satisfactorily for similar reasons to those stated for automating permutation
manipulation. However, some tasks were delegated to custom tactics, for
example instantiating cofinitely quantified hypotheses with a newly introduced
fresh variable.

Initially, an implicit notion of weakening was chosen so as to be in line with
\citeauthor{Wadler:2014}'s definitions. However, explicit weakening is
required to differentiate the principal cut elimination rules for the
interactions between servers and clients (spawning and weakening). The paper
presentation does not specify where weakening could occur; it is tacitly
assumed weakening can always be pushed further up the derivation tree to a
base rule e.g. \coqe$cp_fwd$ rule. Contraction is not supported in the
formalisation due to the difficulty in specifying its principal cut
elimination rule. In order to specify the rule one requires to show the
following:
\begin{gather*}
\tm{\nu A.(!0(A).P \mid ?z[[A]].Q)}
\quad \becomes \quad \\
\tm{\nu A.(!0(A).P \mid \nu A.(!0(A).P \mid \key{contract}_{\Gamma} Q))}
\end{gather*}

@@

Unfortunately, the reduction relation does not have knowledge of the
environment, $\Gamma$, typing process $\tm{P}$ because this information is
only available after the compound cut is destructed, yielding a derivation for
$\tm{P}$. This fact does not impact upon expressing weakening because only the
names (not the types in the bindings) within the environment are required for
weakening.

In the empty choice rule there may only be a single binding in the context
(cf. an arbitrary context in \citeauthor{Wadler:2014}'s
presentation~\cite{Wadler:2014}). \todo{look into a fix for the empty choice
  rule} While this diverges with the presentation of linear logic it became
necessary to do this because otherwise there would be no relationship between
the free variables within a process and its environment. no there is a clear
iff relationship which permits reduction rules to apply such as i think
weakening but it should be any rule which takes some fvs of proc and infers
something about environments from that...otherwise what it is not at all clear
how to encode the reductions in a relation, the subprocesses environments are
created from destructing the environment

@@

\subsubsection{Top-level Cut Elimination}

The theorem for top-level cut elimination can be expressed as:
\begin{coq}
Theorem proc_cut_elim:
  forall P $\Gamma$
         (WT: P $\tpcp$ $\Gamma$),
    exists Q, P $\becomes^\star$ Q /\ $\sim$ is_cut Q.
\end{coq}

where $\becomes^\star$ denotes the reflexive and transitive closure of
$\becomes$ and $\sim$ \coqe$is_cut Q$ expresses that Q is not a cut.

Unfortunately, the theorem has eluded attempts at mechanised proof so far. The
hypothesis of a well-typed process \coqe$P$ is not sufficient to prove the
theorem. For instance, the \coqe$cp_cut$ constructor does not ensure that
there is a duality relationship between the subprocesses. Following informal
proofs, it is likely to require a termination measure involving the length of
the derivation is needed. Additionally, the locally nameless representation
creates subgoals for which it is not clear how to prove. For instance, a
commuting conversion is only defined if the logical operator corresponds to a
free variable but induction on \key{WT} yields two cases one of which applies
the logical operator to the top-level cut variable. Since there is no
relationship between the subprocesses, in general it will not be possible to
apply a principal cut rule.
